---
title: "Project 1 Checkpoint 2"
format: html
author: Braden O'Donnell, Brayden Philabaum, Grayson Six
---

## Load in necessary libraries

```{r, message=FALSE}
library(tidyverse)
library(purrr)
```

## Importing Democracy Data from Github

```{r, message=FALSE}
democracy_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-11-05/democracy_data.csv')
```

# Part 1

## Summary on: What countries have the most regime changes over time?

```{r}
count_regime_changes<-function(country){
  country_data<-democracy_data |> 
    filter(country_name==country)|> # Filters so data is only inputed country
    arrange(year) # Sorts data from earliest year to most recent year
  
regime<-country_data$regime_category_index

# Creates a vector that returns true when regime has been changed from previous year
changes<- regime != lag(regime) 

sum(changes, na.rm=TRUE) # Sum the amount of times the regime has changed

}

results<- democracy_data |>
  distinct(country_name) |> # Keeps only unique country names
  mutate(n_changes=map_dbl(
    country_name,count_regime_changes
    )) # Runs the function on each country and counts the number of changes
  
results |> slice_max(n_changes,n=10)

```

# Part 2

## Summary on: How often do countries have full suffrage when they are democratic?

```{r}
prop_suffrage_dem <- function(country) {
  democracy_data |>
    filter( # Keeps the years where the selected country is a democracy
      country_name == country,
      is_democracy == TRUE 
    ) |>
    # Computes the proportion of democratic years with full suffrage
    summarise(prop_suffrage = mean(has_full_suffrage, na.rm = TRUE)) |>
    # Extracts value so function returns numeric result
    pull(prop_suffrage)
}
 
# Top 10 
democracy_data |>
  distinct(country_name) |>
  mutate(
    prop_suffrage = map_dbl(country_name, prop_suffrage_dem)
  ) |>
  slice_max(prop_suffrage, n = 10)

# Bottom 10
democracy_data |>
  distinct(country_name) |>
  mutate(
    prop_suffrage = map_dbl(country_name, prop_suffrage_dem)
  ) |>
  slice_min(prop_suffrage, n = 10)
  
```

Each number represents the proportion of democractic years where the country had full suffrage.

# Part 3

## Summary on: **Proportion of country years (1950-2020) with a female leader within each regime, country, election system**

```{r}
sum_female_leader <- function(democ_data, columns, group){ 
    
    female_leader_any <- democ_data |> 
        rowwise() |>    # Selects the rows as groups to check over each year
        mutate(female_leader = any(c_across(all_of(columns)), na.rm = TRUE)) |>
                        # Checks across defined columns for a female leader
        ungroup()       # Undoes rowwise() grouping
    
    prop_female <- female_leader_any |> 
        group_by(across(all_of(group))) |>  # Grouped by group = argument
        summarise(prop_female_leader = mean(female_leader, na.rm = TRUE))
                        # mean of TRUE/FALSE for the proportion 

    return(prop_female) # returns the group argument and the proportion
}

sum_female_leader(democracy_data, 
                     columns = c("is_female_monarch", "is_female_president"), 
                     group = "country_name")

sum_female_leader(democracy_data, 
                     columns = c("is_female_monarch", "is_female_president"), 
                     group = "regime_category")

sum_female_leader(democracy_data, 
                     columns = c("is_female_monarch", "is_female_president"), 
                     group = "election_system")
```

Female_leader is a constructed variable for type of leader monarch and president. It is a logical for each row and either there is a female leader or not.

# Part 4

## Summary on: **Which countries have the least stable governments, measured by the frequency of regime changes?**

```{r}
#regime_stability creates a variable avg_reg_change that reports the avg regime changes per yer for that country
regime_stability <- function(country, data = democracy_data){
  result <- data |>
    filter(country_name == country) |>
    arrange(year) |>
    mutate(
      change = regime_category_index != lag(regime_category_index) # creates logical variable change
    ) |>
    summarise(avg_reg_change = mean(change, na.rm = TRUE))
  
  return(result$avg_reg_change)
}

stability_data <- democracy_data |>
  distinct(country_name) |>
  mutate(avg_reg_change = map_dbl(country_name,regime_stability)) |>
  arrange(desc(avg_reg_change))


ggplot(stability_data[1:3, ], 
       aes(x = reorder(country_name, -avg_reg_change), y = avg_reg_change)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
    labs(
    title = "Least Stable Countries by Regime Change Frequency",
    x = "Country",
    y = "Average Proportion of Years with Regime Change"
  ) +
  theme_minimal()
```

Guinea-Bissau, Bangladesh, and Thailand exhibit the highest frequencies of regime change, indicating they are among the least politically stable countries in the data set.
